---
title: "ACS variables"
format:
  html:
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
---


Pilot the calculation of Homeownership (v153), Severe Housing Problems (v136), Some College (v069), High School Completion (v168), Residential Segregation – Black/White (v141), Disconnected Youth (v149), % Not Proficient in English (v059), Children in Single-Parent Households (v082), Driving Alone to Work (v067), Long Commute – Driving Alone (v137) 

# Completion status

##  % Not Proficient in English (v059)
- Diff in CI, only NA vs 0/1?

## Driving Alone to Work (v067)
- Calc by race not implemented

## Some College (v069)
- Diff in CI, only NA vs 0/1?

## Children in Single-Parent Households (v082)
- Complete, numerically equal without zero-max rule

## Severe Housing Problems (v136)
- TODO from scripts/v136_severe_housing_problems.Rmd
- Uses data from HUD, skipping for now.

## Long Commute – Driving Alone (v137)
- Complete, numerically equal without zero-max rule.
- No ratio calculation, proportions already in raw data.
- Wrong denominator value in source code (`denominator = S0802_C01_001E`, total)? Should be `S0802_C02_001E`, drove alone.

##  Residential Segregation – Black/White (v141)
- TODO from scripts/v141_residential_segregation.qmd
- Non-standard ACS, columns from three separate tables (Total, White and Black)
- Aggregation to counties from tracts

## Disconnected Youth (v149)
- Complete, numerically equal without zero-max rule
- Special case handling, maybe include in general function?

## Homeownership (v153)
- Complete, numerically equal

## High School Completion (v168)
- Diff in CI, only NA vs 0/1?


# Libraries

```{r}

library(tidyverse)

source("R/acs.R")

ACS_YEAR <- 2023

```


# Raw data


```{r}

x <- get_raw_table(2023, "B15001")

x %>% head()


```


# v059 - % Not Proficient in English


Numerator:

The numerator is the number of people aged 5 and over who self-reported they are not proficient in speaking English (speak English either “not well” or “not at all”).

Denominator:

The denominator is the total county population aged 5 and over.


```{r}

numerator <- expr(B16005_007E + B16005_012E + B16005_008E + B16005_013E + B16005_017E + B16005_022E + B16005_018E + B16005_023E + B16005_029E + B16005_030E + B16005_034E + B16005_035E + B16005_039E + B16005_040E + B16005_044E + B16005_045E)
denominator <- expr(B16005_001E)

x <- get_raw_table(2023, "B16005") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v059") %>%
  arrange(statecode, countycode)

v059 <- x %>%
  select(statecode, countycode, starts_with("v059"))

```

```{r}
x1 <- read_csv("measure_datasets/v059_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v059, tolerance = 1e-4, list_as_map = TRUE)

```


# v067 - PERCENT OF LABOR FORCE THAT DRIVES ALONE TO WORK

numerator =	number of workers who commute alone to work via car, truck, or van.
denominator	= total workforce
raw_value	= Percentage of workers who drive alone to work

```{r}

numerator <- expr(B08301_003E)
denominator <- expr(B08301_001E)

x <- get_raw_table(2023, "B08301") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v067")

v067 <- x %>%
  select(statecode, countycode, starts_with("v067"))

```

```{r}
x1 <- read_csv("measure_datasets/v067_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v067, tolerance = 1e-4, list_as_map = TRUE)

```


## v067_other - Driving Alone to Work

`race_aian`	AIAN	Percentage of American Indian and Alaska Native workers who drive alone to work	
`race_asian`	Asian	Percentage of Asian/Pacific Islander workers who drive alone to work
`race_black`	Black	Percentage of Black workers who drive alone to work	Percentage
`race_hispanic`	Hispanic	Percentage of Hispanic workers who drive alone to work
`race_white`	white	Percentage of non-Hispanic white workers who drive alone to work


```{r}
var_ls_067_other <- c(
  'B08505B_002E', 'B08505H_002E', 'B08505I_002E', 'B08505D_002E', 'B08505E_002E',
  'B08505C_002E', 'B08505B_001E', 'B08505H_001E', 'B08505I_001E', 'B08505D_001E',
  'B08505E_001E', 'B08505C_001E', 'B08505B_002M', 'B08505H_002M', 'B08505I_002M',
  'B08505D_002M', 'B08505E_002M', 'B08505C_002M', 'B08505B_001M', 'B08505H_001M',
  'B08505I_001M', 'B08505D_001M', 'B08505E_001M', 'B08505C_001M'
)
```


# v069 - SOME COLLEGE

numerator =	Total number of individuals ages 25-44 with at least some post-secondary education
denominator	= Total number of individuals ages 25-44.
raw_value	= Percentage of adults age 25-44 with some post-secondary education

```{r}

numerator <- expr(B15001_015E + B15001_016E + B15001_017E + B15001_018E + B15001_023E + 
        B15001_024E + B15001_025E + B15001_026E + B15001_056E + B15001_057E + B15001_058E +  
        B15001_059E + B15001_064E + B15001_065E + B15001_066E + B15001_067E)
denominator <- expr(B15001_011E + B15001_019E + B15001_052E + B15001_060E)

x <- get_raw_table(2023, "B15001") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v069")

v069 <- x %>%
  select(statecode, countycode, starts_with("v069"))

```


```{r}

x1 <- read_csv("measure_datasets/v069_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v069, tolerance = 1e-4, list_as_map = TRUE)

```

## track differences

```{r}
cmp_vars <- c("statecode", "countycode", "numerator", "denominator", "rawvalue", "cilow", "cihigh")
x0 <- read_csv("measure_datasets/v069_r2025.csv") %>%
  arrange(statecode, countycode) %>%
  rename_with(\(x) str_remove(x, "^v..._")) %>%
  select(all_of(cmp_vars))

numerator <- expr(B15001_015E + B15001_016E + B15001_017E + B15001_018E + B15001_023E + 
        B15001_024E + B15001_025E + B15001_026E + B15001_056E + B15001_057E + B15001_058E +  
        B15001_059E + B15001_064E + B15001_065E + B15001_066E + B15001_067E)
denominator <- expr(B15001_011E + B15001_019E + B15001_052E + B15001_060E)

x1 <- get_raw_table(2023, "B15001") %>%
  standardize_fips() %>% 
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE)

cmp <- cmp_vars[-(1:2)] %>%
  set_names() %>%
  map(\(var) {
    waldo::compare(x0[[var]], x1[[var]], tolerance = 1e-4)
  })

cmp$cilow

almost_equal <- function(x, y, rel_tol = 0.001, abs_tol = Inf) {
  abs_dif <- abs(x - y)
  rel_dif <- ifelse(x == 0, 0, abs_dif / abs((x + y) / 2))
  !is.na(x) & !is.na(y) & abs_dif <= abs_tol & rel_dif <= rel_tol
}

x <- full_join(
  select(x0, statecode, countycode, v0 = cilow),
  select(x1, statecode, countycode, v1 = cilow),
  by = c("statecode", "countycode")
) %>%
  filter(!almost_equal(v0, v1))

x1e <- x1 %>%
  right_join(x) %>%
  select(!starts_with("B15001"))



```

```{r}

```


# v082 - SINGLE-PARENT HOUSEHOLDS

numerator =	Number of children that live in single-parent households
denominator	= Number of children in households
raw_value	= Percentage of children that live in single-parent households


```{r}

numerator <- expr(B09005_004E + B09005_005E)
denominator <- expr(B09005_001E)

x <- get_raw_table(2023, "B09005") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  apply_suppression() %>%
  # repeating original edge-case code, might want to generalize to work with any zero values
  # setting '15005' to missing because numerator and denominator are 0 which isn't helpful
  mutate(across(c(cilow, cihigh), ~if_else(is.nan(rawvalue), NA, .))) %>% 
  mutate(numerator = if_else((statecode=="15" & countycode=="005"), NA, numerator),
         denominator = if_else((statecode=="15" & countycode=="005"), NA, denominator),
         cilow = if_else((statecode=="15" & countycode=="005"), NA, cilow),
         cihigh = if_else((statecode=="15" & countycode=="005"), NA, cihigh)) %>% 
  mutate(sourceflag = NA) %>%
  add_col_prefix("v082")

v082 <- x %>%
  select(statecode, countycode, starts_with("v082"))

```


```{r}

x1 <- read_csv("measure_datasets/v082_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v082, tolerance = 1e-4, list_as_map = TRUE)

```


# v136 - Severe Housing Problems

TODO


# v137 - LONG COMMUTE - DRIVING ALONE 

numerator =	number of workers who drive alone (via car, truck, or van) for more than 30 minutes during their commute
denominator	= number of workers who drive alone (via car, truck, or van) during their commute
raw_value	= Among workers who commute in their car alone, the percentage that commute more than 30 minutes


```{r}
tabid <- "S0802"
cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid))

denominator <- expr(S0802_C02_001E)
rawvalue <- expr(S0802_C02_086E + S0802_C02_087E + S0802_C02_088E + S0802_C02_089E)

x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  # no ratio, using general function to handle variance calc
  # calc_acs_ratio(num_expr = !!rawvalue, den_expr = !!denominator, zero_max = FALSE) %>%
  calc_acs_ratio(num_expr = !!rawvalue, den_expr = !!denominator) %>%
  mutate(
    rawvalue = numerator / 100,
    numerator = NA,
    var = var_num / 10000,
    var_num = NA,
    var_prop = NA,
    moe = sqrt(var) * 1.96,
    cilow = rawvalue - moe, 
    cihigh = rawvalue + moe
  ) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v137")

v137 <- x %>%
  select(statecode, countycode, starts_with("v137"))

```


```{r}

x1 <- read_csv("measure_datasets/v137_r2025.csv") %>%
  arrange(statecode, countycode) %>%
  # old uses wrong denominator
  select(-v137_denominator)

waldo::compare(x1, v137, tolerance = 1e-4, list_as_map = TRUE)

```




## var list
```{r}

```

## process data
```{r}
longcommute <- acs5 %>% 
  select(1:3, all_of(var_ls_137)) %>% 
  rename(denominator = S0802_C01_001E) %>% 
  glimpse()

longcommute_2 <- longcommute %>% 
  mutate(
    numerator = NA_real_,
    rawvalue = S0802_C02_086E + S0802_C02_087E + S0802_C02_088E + 
           S0802_C02_089E,
    se = (sqrt(S0802_C02_086M**2 + S0802_C02_087M**2 + S0802_C02_088M**2 +
                    S0802_C02_089M**2))*1.96/1.645
       ) %>% 
  mutate(cilow = rawvalue - se, 
         cihigh = rawvalue + se,
         sourceflag = NA_real_)

standard_cols <- c('statecode', 'countycode', 'numerator', 'denominator',
                   'rawvalue', 'cilow', 'cihigh', 'sourceflag')

longcommute_3 <- longcommute_2 %>% 
  select(all_of(standard_cols)) %>% 
  mutate(across(c(rawvalue, cilow, cihigh), ~ .x/100) ) %>% 
  glimpse()
```

## suppression + standardize column names

```{r}
longcommute_4 <- longcommute_3 %>%
  apply_suppression() %>% 
  # standard column names
  standardize_col_names("v137") %>% 
  glimpse()
```

## standardize fips and add flag_CT

```{r}
v137 <- longcommute_4 %>% 
  standardize_fips() %>% 
  add_flag_CT(vnum = "v137", old="U", new = "A")

v137
```

## save 

```{r}
v137 %>% 
  save_data("v137")
```


# v141 - Residential Segregation – Black/White
- TODO from scripts/v141_residential_segregation.qmd


# v149 - DISCONNECTED YOUTH

numerator =	number of people, ages 16-19, who are neither working nor in school
denominator	= total county population, ages 16-19
raw_value	= Percentage of teens and young adults ages 16-19 who are neither working nor in school


```{r}

numerator <- expr(B14005_010E + B14005_011E + B14005_014E + B14005_015E +
           B14005_024E + B14005_025E + B14005_028E + B14005_029E)
denominator <- expr(B14005_001E)

x <- get_raw_table(2023, "B14005") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  # could go inside apply_suppresion?
  mutate(
    rawvalue = case_when(
      rawvalue == 0 ~ NA_real_,
      rawvalue == 1 ~ NA_real_,
      # standard error too high
      sqrt(var) / rawvalue > 0.35 ~ NA_real_,
      TRUE ~ rawvalue
    )
  ) %>%
  mutate(across(c(numerator, denominator, cilow, cihigh), 
                ~if_else(is.na(rawvalue), NA_real_, .x))) %>% 
  add_col_prefix("v149")

v149 <- x %>%
  select(statecode, countycode, starts_with("v149"))

```


```{r}

x1 <- read_csv("measure_datasets/v149_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v149, tolerance = 1e-4, list_as_map = TRUE)

```


# v153 - Homeownership

Description: Percentage of owner-occupied housing units.

Numerator: The numerator is the total number of owner-occupied housing units in a county.

Denominator: The denominator is the total occupied housing units in a county.


```{r}
tabid <- "B25003"
cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid))

numerator <- expr(B25003_002E)
denominator <- expr(B25003_001E)

x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v153")

v153 <- x %>%
  select(statecode, countycode, starts_with("v153"))

```


```{r}

x1 <- read_csv("measure_datasets/v153_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v153, tolerance = 1e-4, list_as_map = TRUE)

```


# v168 - HIGH SCHOOL COMPLETION

numerator =	# Completed High School (Adults age 25 and over with a high school diploma or equivalent)
denominator =	Population	(Adults age 25 and over)
raw_value	= % Completed High School


```{r}

numerator <- expr(S1501_C01_014E)
denominator <- expr(S1501_C01_006E)

x <- get_raw_table(2023, "S1501") %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA_real_) %>%
  add_col_prefix("v168")

v168 <- x %>%
  select(statecode, countycode, starts_with("v168"))

```


```{r}

x1 <- read_csv("measure_datasets/v168_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x1, v168, tolerance = 1e-4, list_as_map = TRUE)

```
