---
title: "ACS variables"
format:
  html:
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
---

# top



```{r}
library(logger)
library(tidyverse)


source("R/util.R", local = (util <- new.env()))

log_threshold(DEBUG)


ipath <- list(
  # https://www2.census.gov/programs-surveys/acs/summary_file/2023/table-based-SF/data/5YRData/5YRData.zip
  acs_2023_5_zip = "data/raw/acs_2023_5/5YRData.zip"
)

opath <- list(
)


```


# data

Nation
api.census.gov/data/2022/acs/acs1/profile?get=NAME,DP05_0001E&ucgid=0100000US&key=KEY

All states
api.census.gov/data/2022/acs/acs1/profile?get=NAME,DP05_0001E&ucgid=pseudo(0100000US$0400000)&key=KEY

All counties
api.census.gov/data/2022/acs/acs1/profile?get=NAME,DP05_0001E&ucgid=pseudo(0100000US$0500000)&key=KEY

```{r}
source("R/acs.R")


x <- get_raw_table(2023, "B15001")

glimpse(x)
```



## summary tables

```{r}
table_file_names <- unzip(ipath$acs_2023_5_zip, list = TRUE)
table_file_name <- table_file_names %>%
  filter(str_detect(Name, "b15001")) %>%
  pull(Name)
stopifnot(length(table_file_name) == 1)

df_table <- read_delim(
  unz(ipath$acs_2023_5_zip, table_file_name), 
  delim = "|",
  col_types = list(GEO_ID = "c", .default = "d")
)

df_table %>% glimpse()

```



# v069 - SOME COLLEGE

numerator =	Total number of individuals ages 25-44 with at least some post-secondary education
denominator	= Total number of individuals ages 25-44.
raw_value	= Percentage of adults age 25-44 with some post-secondary education

## var list

```{r}
var_ls_069 <- c(
  'B15001_E015', 'B15001_E016', 'B15001_E017', 'B15001_E018', 'B15001_E023',
  'B15001_E024', 'B15001_E025', 'B15001_E026',
  'B15001_E056', 'B15001_E057', 'B15001_E058', 'B15001_E059', 'B15001_E064',
  'B15001_E065', 'B15001_E066', 'B15001_E067',
  'B15001_E011', 'B15001_E019', 'B15001_E052', 'B15001_E060',
  'B15001_M015', 'B15001_M016', 'B15001_M017', 'B15001_M018', 'B15001_M023',
  'B15001_M024', 'B15001_M025', 'B15001_M026',
  'B15001_M056', 'B15001_M057', 'B15001_M058', 'B15001_M059', 'B15001_M064',
  'B15001_M065', 'B15001_M066', 'B15001_M067',
  'B15001_M011', 'B15001_M019', 'B15001_M052', 'B15001_M060'
)



```






# v168 - HIGH SCHOOL COMPLETION

numerator =	# Completed High School (Adults age 25 and over with a high school diploma or equivalent)
denominator =	Population	(Adults age 25 and over)
raw_value	= % Completed High School

## var list
```{r}
var_ls_168 <- c('S1501_C01_014E', 'S1501_C01_014M',
                'S1501_C01_006E', 'S1501_C01_006M')
```


```{r}
table_file_names %>%
  pull(Name) %>%
  basename() %>%
  str_extract("-(.+)\\.dat", group = 1) %>%
  str_sub(1, 2) %>% table()
```

## process data
```{r}
hs_completion <- acs5 %>% 
  select(1:3, all_of(var_ls_168)) %>% 
  rename(numerator = S1501_C01_014E, 
         denominator = S1501_C01_006E) %>% 
  mutate(se_num_square = S1501_C01_014M**2,
         se_denom_square = S1501_C01_006M**2) %>% 
  glimpse()

hs_completion_2 <- hs_completion %>% 
  compute_se_ci()

standard_cols <- c('statecode', 'countycode', 'numerator', 'denominator',
                   'rawvalue', 'cilow', 'cihigh', 'sourceflag')

hs_completion_3 <- hs_completion_2 %>% 
  select(all_of(standard_cols))
```


# tidycensus test

```{r}
.libPaths()

```

```{r}

library(tidycensus)




```


```{r}
x <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "VT", 
              year = 2024)
x

```


```{r}
acs5_2024 <- load_variables(2024, "acs5", cache = TRUE)
acs5s_2024 <- load_variables(2024, "acs5/subject", cache = TRUE)

```


```{r}
var_ls_166 <- c('S2801_C01_014E', 'S2801_C01_014M',
                'S2801_C01_001E', 'S2801_C01_001M')

x <- get_acs(geography = "county", 
              variables = var_ls_166, 
              state = "VT", 
              year = 2024)         

acs5s_2024
x %>%
  left_join(acs5s_2024, join_by(variable == name))

```


# Census API


```{r}


```


# v082	- Children in single-parent households	

```{r}

library(tidyverse)

compute_se_ci <- function(df){
  df %>% 
    mutate(rawvalue = numerator/denominator,
       se_num = (sqrt(se_num_square))/1.645,
       se_denom = (sqrt(se_denom_square))/1.645) %>% 
  mutate(se_test = se_num^2 - (rawvalue^2 * se_denom^2)) %>% 
  mutate(se_pct = case_when(
    se_test>=0 ~ sqrt(se_num**2 - (rawvalue^2 * se_denom^2))/denominator,
    se_test< 0 ~ sqrt(se_num**2 + (rawvalue^2 * se_denom^2))/denominator
  )) %>% 
  mutate(cilow = rawvalue - se_pct*1.96, 
         cihigh = rawvalue + se_pct*1.96,
         sourceflag = NA_real_)
}

apply_suppression <-  function(df){
  df %>% 
  mutate(
    cilow = case_when(cilow < 0 ~ 0,
                      is.na(cihigh) ~ NA_real_,
                      TRUE ~ cilow),
    cihigh = case_when(cihigh > 1 ~  1, 
                       is.na(cilow) ~ NA_real_,
                       TRUE ~ cihigh)
  )
}

standard_cols <- c('statecode', 'countycode', 'numerator', 'denominator',
                   'rawvalue', 'cilow', 'cihigh', 'sourceflag')


standardize_col_names <- function(df, col_prefix){
  res <- df %>% 
    rename(
           !!(paste0(col_prefix, "_rawvalue")) := rawvalue, 
           !!(paste0(col_prefix, "_numerator")) := numerator,
           !!(paste0(col_prefix, "_denominator")) := denominator,
           !!(paste0(col_prefix, "_cilow")) := cilow,
           !!(paste0(col_prefix, "_cihigh")) := cihigh,
           !!(paste0(col_prefix, "_sourceflag")) := sourceflag
           )
  return(res)
}

var_ls_082 <- c('B09005_001E', 'B09005_004E', 'B09005_005E', 
                'B09005_001M', 'B09005_004M', 'B09005_005M')

proc_wide <- function(wide_df) {
  wide_df %>%
    mutate(numerator = B09005_004E + B09005_005E, 
          denominator = B09005_001E,
          se_num_square = B09005_004M**2 + B09005_005M**2,
          se_denom_square = B09005_001M**2) %>% 
    compute_se_ci() %>%
    apply_suppression() %>%
    select(all_of(standard_cols)) %>%
    standardize_col_names("v082")
}

v082 <- list()

```


## CHRR

```{r}
v082$chrr %>%
  filter(countycode == "000", statecode != "00") %>%
  tail()
```

```{r}

v082$chrr <- read_csv("measure_datasets/v082_r2025.csv")

```

## tidycensus


```{r}

library(tidycensus)

raw_tc <- get_acs(geography = "county", variables = var_ls_082, year = 2023)

wide_tc <- raw_tc %>%
  mutate(statecode = str_sub(GEOID, 1, 2), countycode = str_sub(GEOID, 3, 5)) %>%
  select(statecode, countycode, variable, E = estimate, M = moe) %>%
  pivot_longer(c(E, M)) %>%
  mutate(name = paste0(variable, name)) %>%
  pivot_wider(id_cols = c(statecode, countycode))


v082$tc <- wide_tc %>%
  proc_wide()

```


## Census API

```{r}
library(httr2)

url <- "https://api.census.gov/data/2023/acs/acs5?get=group(B09005)&for=county:*"

raw_api <- request(url) %>%
  req_perform() %>%
  resp_body_json(simplifyVector = FALSE)

wide_api <- raw_api[-1] %>%
  lapply(\(x) setNames(x, raw[[1]])) %>%
  bind_rows() %>%
  select(statecode = state, countycode = county, all_of(var_ls_082)) %>%
  type_convert()

v082$api <- wide_api %>%
  proc_wide()


```

## FTP summary file


```{r}
raw_ftp <- util$get_file(
    "data/raw/acsdt5y2023-b09005.dat",
    "https://www2.census.gov/programs-surveys/acs/summary_file/2023/table-based-SF/data/5YRData/acsdt5y2023-b09005.dat"
  ) %>%
  read_delim(
    delim = "|",
    col_types = list(GEO_ID = "c", .default = "d")
  )

wide_ftp <- raw_ftp %>%
  filter(str_sub(GEO_ID, 1, 3) == "050") %>%
  mutate(statecode = str_sub(GEO_ID, 10, 11), countycode = str_sub(GEO_ID, 12, 14)) %>%
  # B09005_E001 -> B09005_001E
  rename_with(
    \(x) paste0(substr(x, 1, 7), substr(x, 9, 11), substr(x, 8, 8)),
    .cols = starts_with("B09005")) %>%
  select(statecode, countycode, all_of(var_ls_082))

v082$ftp <- wide_ftp %>%
  proc_wide()

```

## compare

```{r}
v082_df <- v082 %>%
  bind_rows(.id = "method") %>%
  filter(countycode != "000")
  
v082_df %>%
  arrange(statecode, countycode, method)

v082_var <- v082_df %>%
  summarize(across(where(is.numeric), \(x) max(x) - min(x)), .by = c(statecode, countycode))
  
v082_var %>%
  skimr::skim()

# sus
v082_var %>%
  filter(if_any(where(is.numeric), \(x) x > 0.1)) %>%
  select(statecode, countycode) %>%
  left_join(v082_df)


```


# formula


```{r}

```

