---
title: "v141 residential segregation"
author: "how"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

-   Index of dissimilarity where higher values indicate greater residential segregation between Black and White county residents.

## Raw data access

We access data using the tidycensus R package. Data come from ACS tables B11002 (all), B11002A (white), and B11002B (black) at the census tract level of the 2019-2023 ACS 5-year estimates. These data are accessible here: https://data.census.gov/table/ACSDT1Y2022.B11002

# import packages

```{r}
library(tidyverse)
library(haven)
library(tidycensus)
```

# standard county and state fips

-   The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat"),
  read_sas( "../inputs/state_fips.sas7bdat")) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# load data using the tidycensus package

```{r}
# Variables
vars <- c(
  "B11002_001E",   # Total households
  "B11002A_001E",  # White alone, not Hispanic
  "B11002B_001E"   # Black or African American alone
)

states <- unique(fips_codes$state)[1:51]  # state abbreviations from tidycensus (DC included)

tract_data <- map_dfr(states, function(st) {
  get_acs(
    geography = "tract",
    variables = vars,
    year = 2023,
    survey = "acs5",
    output = "wide",
    state = st
  )
})


tract_data = tract_data %>% 
  mutate(statecode = substr(GEOID, 1,2), 
         countycode = substr(GEOID, 3,5), 
         bi = B11002B_001E, #tract level black pop
         wi = B11002A_001E, #tract level white pop 
         toti = B11002_001E) 
```

calculate county level totals (we do this instead of pulling county level data directly from ACS since we found a discrepancy between county and tract data in 2025)

```{r}
county_level = tract_data %>% group_by(statecode, countycode) %>% 
  summarize(btot = sum(bi, na.rm = TRUE), 
            wtot = sum(wi, na.rm = TRUE), 
            tot = sum(toti, na.rm = TRUE))
```

add the ct crosswalk (for 2025 only)

```{r}
#load the ct tract crosswalk 
cttract = haven::read_sas(here("inputs/ct_tract_crosswalk.sas7bdat"))

ctonly = tract_data %>% filter(statecode == "09") %>% 
  mutate(GEOID = stringr::str_pad(GEOID, width =11,side = "left", pad = "0"))

ctwalk = merge(cttract, ctonly, by.x = "tract_fipscode_2022",
               by.y = "GEOID")

ctwalk = ctwalk %>% select(statecode, countycode, bi, wi, toti) %>% 
  group_by(countycode) %>% 
  mutate(btot = sum(bi, na.rm = TRUE), 
         wtot = sum(wi, na.rm = TRUE))

#ctwalk now contains the corrected ct fipscodes for 2025 release 


```

put all the data together

```{r}
data <- tract_data %>% 
  left_join(
    county_level %>% select(statecode, countycode, btot, wtot),
    by = c("statecode", "countycode")
  ) 


data_ct = data %>% filter(statecode != "09") %>% 
  bind_rows(ctwalk)

# check that there are no geoids in `data` but not in `data_ct`
geoids_not_in_data_ct <- data_ct %>%
  anti_join(data, by = "GEOID") %>%
  select(GEOID) %>%
  distinct()
```

# calculate the index of dissimilarity for each state

```{r}

# get state totals first 
# note that no suppression is applied!! 
s1 = data_ct %>% 
  filter(countycode != "000") %>% 
  group_by(statecode) %>% 
  summarize(
    B = sum(bi),  # Total black population in state
    W = sum(wi), # Total white population in state
    bi_B = bi / B,  # Proportion of black population in tract
    wi_W = wi / W,  # Proportion of white population in tract
    abs_diff = abs(bi_B - wi_W))  # Set to NA if missing

states = s1 %>% group_by(statecode) %>% 
  summarize(
    rawvalue = 50 * sum(abs_diff, na.rm = TRUE))  # (1/2) SUM|(bi/B) - (wi/W)| * 100)
states$countycode = "000"

```

# Calculate the index of dissimilarity components for each tract for countycalcs

```{r}

c1 <- data_ct %>% filter(countycode != "000") %>% 
  mutate(
    B = btot,  # Total black population in county
    W = wtot,  # Total white population in county
    bi_B = bi / B,  # Proportion of black population in tract
    wi_W = wi / W,  # Proportion of white population in tract
    abs_diff = if_else(is.na(bi_B) | is.na(wi_W), NA_real_, abs(bi_B - wi_W))  # Set to NA if missing
  )


# Sum the components by county and calculate the index
county <- c1 %>%
  group_by(statecode, countycode) %>% 
  filter(B>=100) %>% 
  summarize(
   rawvalue = 50 * sum(abs_diff))  # (1/2) SUM|(bi/B) - (wi/W)| * 100)

county$rawvalue = ifelse(county$rawvalue == 0, NA, county$rawvalue)



```

# put state and county together and add the fipscodes back in

```{r}
sctot = dplyr::bind_rows(states, county) 

scall = merge(fips, sctot, by= c("statecode", "countycode"), all.x = TRUE)

scall <- scall %>%
  mutate(
    # Calculate the median for countycode = 000
    rawvalue = if_else((statecode == "00" & countycode == "000"),
      median(rawvalue[countycode == "000"], na.rm = TRUE),
      rawvalue
    )
  )


```

# suppress ulster and suffolk counties due to issues with these data in 2025:

fipscodes 36103 and 36111

```{r}
scall = scall %>% mutate(
  v141_rawvalue = if_else((statecode == "36" & countycode %in% c("103", "111")), 
                     NA, rawvalue)
)
```

# save data

```{r}
save_data(scall, "v141")
```
