---
title: "UW S3 Storage test"
format:
  html:
    author: Anton Babkin
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
    df-print: kable
---

# Libraries and S3 secrets


```{r}
library(tidyverse)
library(arrow)
library(aws.s3)

```


```{r}
Sys.setenv(
  AWS_S3_ENDPOINT = "s3.wisc.edu",
  AWS_DEFAULT_REGION = "campus"
)

Sys.getenv(c("AWS_ACCESS_KEY_ID", "AWS_SECRET_ACCESS_KEY"))
```

# Campus pool

Need to be on campus network or VPN to access.

```{r}
bucketlist()

get_bucket(bucket = "babkin-bucket-01")

get_object("hello.txt", bucket = "babkin-bucket-01") %>%
  rawToChar()

put_object("create new text object from R", object = "hello_r.txt", bucket = "babkin-bucket-01")

"https://campus.s3.wisc.edu/babkin-bucket-01/hello.txt" %>%
  httr2::request() %>%
  httr2::req_perform() %>%
  httr2::resp_body_string()

```

# Web pool

Can be outside of campus network.

```{r}
Sys.setenv(AWS_DEFAULT_REGION = "web")
bucketlist()

get_bucket("babkin-test-02-web")

put_object(
  what = "create new text object from R in Web pool bucket",
  object = "hello_r_web.txt",
  bucket = "babkin-test-02-web",
  acl = "public-read"
)

get_object("hello_r_web.txt", bucket = "babkin-test-02-web") %>%
  rawToChar()

get_acl("hello_r_web.txt", bucket = "babkin-test-02-web") %>%
  xml2::read_xml() %>%
  xml2::as_list()

# DO NOT RUN: put_acl without body argument re-PUTs the object, overwriting it's content with empty string
# put_acl("hello_r_web.txt", bucket = "babkin-test-02-web", acl = "public-read")

```


# Caching of CHRR data

New buckets can not be created programmatically with `aws.s3::put_bucket()`.
Instead, create them in Cyberduck [docs](https://kb.wisc.edu/researchdata/134396#newBucket).


```{r}
Sys.setenv(AWS_DEFAULT_REGION = "web")

bucket_name <- "chrr-s3-demo"
stopifnot(bucket_name %in% bucketlist()$Bucket)

get_bucket(bucket_name)

```


# Read parquet files directly from S3

Upload example files: tracts for racial segregation variable (v141).

```{r}
tabid <- c("B11002A", "B11002B")
file_path <- paste0("raw_data/ACS/2024/", tabid, "_tract.pq") %>%
  set_names(c("w", "b"))
file_path %>%
  map_lgl(\(fp) {
    # upload if does not exist
    object_exists(fp, bucket_name) ||
      put_object(
        object = fp, # destination path in the bucket
        file = fp, # local source file path
        bucket = bucket_name,
        acl = "public-read" # make public for credential-free access
      )
  })

```

S3 storage can be declared as `arrow::S3FileSystem`, and objects within it can be accessed with standard `arrow::open_dataset() ... collect()` workflow.
Entire files are not downloaded to local disk, instead, only a requested subset is retreived.
This is not very relevant for 5Mb files, but can be very efficient for larger datasets.

```{r}

fs <- S3FileSystem$create(
  access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
  secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY"),
  scheme = "https",
  endpoint_override = "web.s3.wisc.edu"
)

ds <- c("w", "b") %>%
  set_names() %>%
  map(\(race) {
    open_dataset(
      sources = file.path(bucket_name, file_path[race]),
      filesystem = fs,
      format = "parquet"
    )
  })

# query data tables stored in S3 without downloading them to local disk
x <- full_join(
    ds$w %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(NAME, ucgid, wht = B11002A_001E) %>%
      collect(),
    ds$b %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(ucgid, blk = B11002B_001E) %>%
      collect(),
    by = "ucgid"
  )
  
# dissimilarity index calculator from the v141 notebook
calc_diss_index <- function(df) {
  df %>%
    # suppress counties with less than 100 black population
    filter(sum(blk) >= 100) %>%
    mutate(
      gr_wht = sum(wht), # total white population in geographic group
      gr_blk = sum(blk), # total black population in geographic group
      tr_diss = abs(wht / gr_wht - blk / gr_blk) # dissimilarity in tract shares of white and black 
    ) %>%
    # index = (1/2) SUM|(b/B) - (w/W)| * 100)
    summarize(
      diss_index = 50 * sum(tr_diss, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(diss_index = replace_values(diss_index, 0 ~ NA))
}

# calculate for Dane, WI dataset queried from S3
x %>%
  calc_diss_index()

# compare with previously calculated value in Dane, WI
read_csv("measure_datasets/v141_r2026.csv") %>%
  filter(fipscode == "55025")

```


# Access public S3 objects without credentials


```{r}

fs_pub <- S3FileSystem$create(
  anonymous = TRUE,
  scheme = "https",
  endpoint_override = "web.s3.wisc.edu"
)

ds_pub <- c("w", "b") %>%
  set_names() %>%
  map(\(race) {
    open_dataset(
      sources = file.path(bucket_name, file_path[race]),
      filesystem = fs_pub,
      format = "parquet"
    )
  })

ds_pub$b %>%
  head() %>%
  select(ucgid, NAME, ends_with("E")) %>%
  collect()

```