---
title: "UW S3 Storage test"
format:
  html:
    author: Anton Babkin
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
    df-print: kable
---

# Libraries and S3 secrets


```{r}
library(tidyverse)
library(arrow)
library(aws.s3)

```


```{r}
Sys.setenv(
  AWS_S3_ENDPOINT = "s3.wisc.edu",
  AWS_DEFAULT_REGION = "campus"
)

Sys.getenv(c("AWS_ACCESS_KEY_ID", "AWS_SECRET_ACCESS_KEY"))
```

# Campus pool

Need to be on campus network or VPN to access.

```{r}
bucketlist()

get_bucket(bucket = "babkin-bucket-01")

get_object("hello.txt", bucket = "babkin-bucket-01") %>%
  rawToChar()

put_object("create new text object from R", object = "hello_r.txt", bucket = "babkin-bucket-01")

"https://campus.s3.wisc.edu/babkin-bucket-01/hello.txt" %>%
  httr2::request() %>%
  httr2::req_perform() %>%
  httr2::resp_body_string()

```

# Web pool

Can be outside of campus network.

```{r}
Sys.setenv(AWS_DEFAULT_REGION = "web")
bucketlist()

get_bucket("babkin-test-02-web")

put_object(
  what = "create new text object from R in Web pool bucket",
  object = "hello_r_web.txt",
  bucket = "babkin-test-02-web",
  acl = "public-read"
)

get_object("hello_r_web.txt", bucket = "babkin-test-02-web") %>%
  rawToChar()

get_acl("hello_r_web.txt", bucket = "babkin-test-02-web") %>%
  xml2::read_xml() %>%
  xml2::as_list()

# DO NOT RUN: put_acl without body argument re-PUTs the object, overwriting it's content with empty string
# put_acl("hello_r_web.txt", bucket = "babkin-test-02-web", acl = "public-read")

```


# Caching of CHRR data

New buckets can not be created programmatically with `aws.s3::put_bucket()`.
Instead, create them in Cyberduck [docs](https://kb.wisc.edu/researchdata/134396#newBucket).


```{r}
Sys.setenv(AWS_DEFAULT_REGION = "web")

bucket_name <- "chrr-s3-demo"
stopifnot(bucket_name %in% bucketlist()$Bucket)

get_bucket(bucket_name)

```


# Read parquet files directly from S3

Upload example files: tracts for racial segregation variable (v141).

```{r}
tabid <- c("B11002A", "B11002B")
file_path <- paste0("raw_data/ACS/2024/", tabid, "_tract.pq") %>%
  set_names(c("w", "b"))
file_path %>%
  map_lgl(\(fp) {
    # upload if does not exist
    object_exists(fp, bucket_name) ||
      put_object(
        object = fp, # destination path in the bucket
        file = fp, # local source file path
        bucket = bucket_name,
        acl = "public-read" # make public for credential-free access
      )
  })

```

S3 storage can be declared as `arrow::S3FileSystem`, and objects within it can be accessed with standard `arrow::open_dataset() ... collect()` workflow.
Entire files are not downloaded to local disk, instead, only a requested subset is retreived.
This is not very relevant for 5Mb files, but can be very efficient for larger datasets.

```{r}

fs <- S3FileSystem$create(
  access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
  secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY"),
  scheme = "https",
  endpoint_override = "web.s3.wisc.edu"
)

ds <- c("w", "b") %>%
  set_names() %>%
  map(\(race) {
    open_dataset(
      sources = file.path(bucket_name, file_path[race]),
      filesystem = fs,
      format = "parquet"
    )
  })

# query data tables stored in S3 without downloading them to local disk
x <- full_join(
    ds$w %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(NAME, ucgid, wht = B11002A_001E) %>%
      collect(),
    ds$b %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(ucgid, blk = B11002B_001E) %>%
      collect(),
    by = "ucgid"
  )
  
# dissimilarity index calculator from the v141 notebook
calc_diss_index <- function(df) {
  df %>%
    # suppress counties with less than 100 black population
    filter(sum(blk) >= 100) %>%
    mutate(
      gr_wht = sum(wht), # total white population in geographic group
      gr_blk = sum(blk), # total black population in geographic group
      tr_diss = abs(wht / gr_wht - blk / gr_blk) # dissimilarity in tract shares of white and black 
    ) %>%
    # index = (1/2) SUM|(b/B) - (w/W)| * 100)
    summarize(
      diss_index = 50 * sum(tr_diss, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(diss_index = replace_values(diss_index, 0 ~ NA))
}

# calculate for Dane, WI dataset queried from S3
x %>%
  calc_diss_index()

# compare with previously calculated value in Dane, WI
read_csv("measure_datasets/v141_r2026.csv") %>%
  filter(fipscode == "55025")

```


# Access public S3 objects without credentials


```{r}

fs_pub <- S3FileSystem$create(
  anonymous = TRUE,
  scheme = "https",
  endpoint_override = "web.s3.wisc.edu"
)

ds_pub <- c("w", "b") %>%
  set_names() %>%
  map(\(race) {
    open_dataset(
      sources = file.path(bucket_name, file_path[race]),
      filesystem = fs_pub,
      format = "parquet"
    )
  })

ds_pub$b %>%
  head() %>%
  select(ucgid, NAME, ends_with("E")) %>%
  collect()

```


# Bigger file performance

In this experiment I compare the time it takes to query a (2875, 3) subset from a synthetic (2039111, 51) test parquet dataset with two methods:

1. Download the file and query from disk.
2. Query directly from S3 using arrow filesystem.

The second method should obviously be faster, and this experiment demonstrates the difference.
Every run is repeated 10 times.

Results:

1. Full download mean (median) time 35 (32) seconds.
2. S3 query mean (median) time 4 (2.3) seconds.


```{r}
bucket_name <- "chrr-s3-demo"
bigger_file <- "temp/bigger.pq"

# repeat the dataframe to make about 2M rows, 100Mb parquet file
x <- read_parquet("raw_data/ACS/2024/B11002A_tract.pq")
x <- (1:23) %>%
  map(\(x1) x) %>%
  bind_rows()
dim(x)

dir.create("temp")
write_parquet(x, bigger_file)
file.size(bigger_file)

put_object(object = bigger_file, file = bigger_file, bucket = bucket_name)

get_bucket_df(bucket_name) %>%
  select(Key, Size)

```


## download + query

```
   user.self sys.self elapsed
       <dbl>    <dbl>   <dbl>
 1      1.75    0.387    9.63
 2      2.70    1.28    55.0 
 3      2.35    1.09    39.3 
 4      2.09    0.802   26.5 
 5      2.08    0.765   24.6 
 6      2.29    0.900   29.0 
 7      2.25    1.08    31.2 
 8      2.32    0.923   33.0 
 9      2.70    1.51    64.5 
10      2.40    1.19    41.3 
```

```{r}
t_local <- (1:10) %>%
  map(\(x) system.time({
    local_file <- tempfile(fileext = ".pq")

    save_object(bigger_file, bucket_name, file = local_file, show_progress = TRUE)
    if (file.size(local_file) != file.size(bigger_file)) 
      warning("downloaded file wrong size ", file.size(local_file))

    x_local <- local_file %>%
      open_dataset(format = "parquet") %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(NAME, ucgid, wht = B11002A_001E) %>%
      collect()
  })) %>%
  bind_rows() %>%
  # strip column attributes
  mutate(across(everything(), c))

t_local
t_local %>% summary()
```


## query from S3

```
   user.self sys.self elapsed
       <dbl>    <dbl>   <dbl>
 1     0.685   0.164     2.49
 2     0.670   0.0900    2.20
 3     0.671   0.0640    2.60
 4     0.692   0.0600    2.33
 5     0.642   0.0440    2.34
 6     0.642   0.0350    2.14
 7     0.858   0.123    11.0 
 8     0.891   0.125    10.8 
 9     0.686   0.0350    2.15
10     0.641   0.0450    1.99
```

```{r}

t_remote <- (1:10) %>%
  map(\(x) system.time({
    fs <- S3FileSystem$create(
      access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
      secret_key = Sys.getenv("AWS_SECRET_ACCESS_KEY"),
      scheme = "https",
      endpoint_override = "web.s3.wisc.edu")

    x_remote <- file.path(bucket_name, bigger_file) %>%
      open_dataset(filesystem = fs, format = "parquet") %>%
      filter(str_starts(ucgid, "1400000US55025")) %>%
      select(NAME, ucgid, wht = B11002A_001E) %>%
      collect()
  })) %>%
  bind_rows() %>%
  # strip column attributes
  mutate(across(everything(), c))

t_remote
t_remote %>% summary()
```